---
title: "CUNY SPS DATA 622 - Machine Learning and Big Data"
subtitle: 'Spring 2021 - Group 3 - Homework 4'
author: "Maryluz Cruz, Samantha Deokinanan, Amber Ferger, Tony Mei, and Charlie Rosemond"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    theme: cerulean
    highlight: pygments
    toc: 3
urlcolor: purple
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

### R Packages

The `R` language is used to facilitate data modeling. The main `R` packages used for data wrangling, visualization, and graphics are listed below.

```{r libraries, echo=TRUE}
# Required R packages
library(tidyverse)
library(kableExtra)
library(skimr)
library(GGally)
library(corrplot)
library(e1071)
library(dummies)
library(caret)

```

### Overview {.tabset .tabset-fade .tabset.-pills}

This analysis uses clustering, principal component analysis (PCA), and the support vector machine (SVM) algorithm to derive insights from a mental health dataset. It begins with exploratory data analysis (EDA) and preprocessing followed by clustering and PCA. It finishes with SVM modeling of demographic and mental-health-related features to predict whether a patient attempted suicide.

All references and a technical appendix of all R code are available at the end of this report.

***
<center> **PROJECT SECTIONS** </center>
***

### Exploratory Data Analysis (EDA)

EDA seeks to understand the data and its nuances. It draws on a combination of summary statistics and univariate and bivariate visualizations to summarize the dataset, its response, and the initial group of possible features. This information will inform preprocessing of the dataset prior to modeling.

#### Dataset

Sourced from an actual research project and deidentified, the dataset contains individual samples describing patients' experiences with ADHD, mood disorders, substance use and misuse, and related mental health behaviors. It consists of 175 samples and 54 columns, including an identifier `Initial` to be dropped and a response `Suicide` to be modeled via SVM. The remaining features for modeling range from demographic information (e.g., `Age`) and various questionnaire responses (e.g., `ADHD.Q1`) to used/abused substances (e.g., `THC`) and medications (e.g., `Psych.meds`). All columns in the dataset are listed and described below.

*Mental Health Dataset Column Definition*
Columns | Variable | Description
---|---|-----
C |  Sex | Male-1, Female-2
D | Race | White-1, African American-2, Hispanic-3, Asian-4, Native American-5, Other or missing data -6
E - W | ADHD self-report scale |  Never-0, rarely-1, sometimes-2, often-3, very often-4
X - AM | Mood disorder questions |  No-0, yes-1; question 3: no problem-0, minor-1, moderate-2, serious-3
AN - AS | Individual substances misuse |   no use-0, use-1, abuse-2, dependence-3
AT | Court Order |   No-0, Yes-1
AU | | Education |  1-12 grade, 13+ college
AV | History of Violence |  No-0, Yes-1
AW | Disorderly Conduct |  No-0, Yes-1
AX | Suicide attempt |  No-0, Yes-1
AY | Abuse Hx |  No-0, Physical (P)-1, Sexual (S)-2, Emotional (E)-3, P&S-4, P&E-5, S&E-6, P&S&E-7
AZ | Non-substance-related Dx |  0 - none; 1 - one; 2 - More than one
BA | Substance-related Dx |  0 - none; 1 - one Substance-related; 2 - two; 3 - three or more
BB | Psychiatric Meds |  0 - none; 1 - one psychotropic med; 2 - more than one psychotropic med

```{r load}
df <- read_csv("https://raw.githubusercontent.com/greeneyefirefly/DATA622-Group3/main/Project_4/ADHD_data.csv")
df <- df %>% rename_all(make.names)
```

```{r summarystats}
skim(df) %>% 
  dplyr::select(-n_missing, -numeric.sd, -numeric.p25, -numeric.p75)
```

Summary statistics reveal useful information to inform pre-processing and modeling. First, the demographic and questionnaire features are entirely complete, though missingness is present elsewhere. Notably, SVM response `Suicide` is approximately 93 percent complete, and `Psych.meds.` is only approximately 33 percent complete. Second, the feature distributions could require transformation prior to modeling. Some of the questionnaire features show clear skewness (e.g., `ADHD.Q12`) or imbalance (e.g., `MD.Q1g`), while many of the substance use/abuse features appear heavily imbalanced towards "no use". Here again, response `Suicide` is imbalanced towards "no attempt". And third, feature `ADHD.Q5` has at least one value of "5", which, per documentation, appears to be an error. This feature value is changed to missing.

```{r data}
#df %>% filter(ADHD.Q5 == 5) # 1 sample
df$ADHD.Q5[df$ADHD.Q5 == 5] <- NA
```

Beyond the to-be-dropped `Initial`, all dataset features are loaded in numeric format, though many of them are categorical if not ordinal in nature. These features thus require conversion to [ordinal] factors. Missing values "NA" are treated as a level, and where applicable, ordered lowest.

```{r initproc}
df[,c(3,4,24:37,46,48:51)] <- lapply(df[,c(3,4,24:37,46,48:51)], function(x) factor(x, ordered=FALSE))
df[,c(5:22,38,40:45,47,52:54)] <- lapply(df[,c(5:22,38,40:45,47,52:54)], function(x) factor(x, ordered=TRUE, exclude=NULL))
df[,c(5:22,38,40:45,47,52:54)] <- lapply(df[,c(5:22,38,40:45,47,52:54)], function(x) fct_relevel(x, NA, after = 0))
df <- df %>% select(-Initial)
```

Visual EDA begins with a focus on the SVM modeling response `Suicide` and its relationships with the demographic and education features as well as the categorical `Abuse`.

`Sex`: For this dataset, the proportion of females having attempted suicide (~0.39) is nearly double the same proportion among males (~0.23).

```{r suicide_sex, fig.height=3}
tab <- with(df, table(Sex, Suicide))
tab <- as.data.frame(prop.table(tab, margin = 1)) %>%
  filter(Suicide == '1')

tab %>%
  mutate(Sex = fct_recode(Sex, Male='1', Female='2')) %>%
  ggplot() +  
  geom_col(aes(x=Sex, y=Freq, fill=Sex)) +
  geom_label(aes(x=Sex, y=Freq, label = round(Freq,2))) +
  theme(legend.position='none') +
  labs(title = 'Proportion attempting suicide, by sex',
       y = 'Proportion')
```

`Race`: For this dataset, approximately half of individuals indicating race/ethnicity as "Other/Missing" have attempted suicide (~0.50). That proportion compares with approximately 0.41 among "White" individuals and approximately 0.22 among "Black" individuals. Zero "Hispanic" individuals in this dataset have attempted suicide.

```{r suicide_race, fig.height=3}
tab <- with(df, table(Race, Suicide))
tab <- as.data.frame(prop.table(tab, margin = 1)) %>%
  filter(Suicide == '1')

tab %>%
  mutate(Race = fct_recode(Race,
                           White='1',
                           Black='2',
                           Hispanic='3',
                           Asian='4',
                           'Native American'='5',
                           'Other/Missing' = '6')) %>%
  ggplot() +  
  geom_col(aes(x=Race, y=Freq, fill=Race)) +
  geom_label(aes(x=Race, y=Freq, label = round(Freq,2))) +
  theme(legend.position='none') +
  labs(title = 'Proportion attempting suicide, by race/ethnicity',
       y = 'Proportion')
```

`Education`: For this dataset, zero individuals with more than fifteen years of formal education have attempted suicide. Level "15", with a proportion having attempted suicide of approximately 1.00, is an outlier relative to lower levels of education, which all show proportions of approximately 0.50 or below.

```{r suicide_education, fig.height=3}
tab <- with(df, table(Education, Suicide))
tab <- as.data.frame(prop.table(tab, margin = 1)) %>%
  filter(Suicide == '1')

tab %>%
  ggplot() +  
  geom_col(aes(x=Education, y=Freq)) +
  theme(legend.position='none') +
  labs(title = 'Proportion attempting suicide, by education level',
       y = 'Proportion')
```

`Abuse`: For this dataset, individuals having been sexually and/or emotionally abused are particularly likely to have attempted suicide. Approximately three quarters of individuals in each of the "Emotional (E)" and "Physical & Sexual & Emotional" categories have attempted suicide. These proportions compare to roughly one in five individuals in each of the "No" and "Physical & Emotional" categories.

```{r suicide_abuse, fig.height=3}
tab <- with(df, table(Abuse, Suicide))
tab <- as.data.frame(prop.table(tab, margin = 1)) %>%
  filter(Suicide == '1')

tab %>%
  mutate(Abuse = fct_recode(Abuse,
                         No='0',
                         'Physical (P)'='1',
                         'Sexual (S)'='2',
                         'Emotional (E)'='3',
                         'P & S'='4',
                         'P & E'='5',
                         'S & E'='6',
                         'P & S & E'='7')) %>%
  ggplot() +  
  geom_col(aes(x=Abuse, y=Freq, fill=Abuse)) +
  geom_label(aes(x=Abuse, y=Freq, label = round(Freq,2))) +
  theme(legend.position='none') +
  labs(title = 'Proportion attempting suicide, by abuse category',
       y = 'Proportion') +
  scale_x_discrete(guide = guide_axis(angle = 45)) 
```

The next group of features for analysis are the numerics--`Age`, `ADHD.Total`, and `MD.TOTAL`--with each one is related to `Suicide` using distribution plots and correlations. For clarity, the plots below do not account for missing values of `Suicide`, which were included in an initial run of plots but whose relationships did not appear to differ in nature from those of the other levels.

```{r suicide_numerics, fig.height=5.5}
df %>%
  filter(!is.na(Suicide)) %>%
  mutate(Suicide = fct_recode(Suicide,
                     'No attempt'='0',
                     Attempt='1')) %>%
  ggpairs(columns = c('Suicide','Age','ADHD.Total','MD.TOTAL'), title = "Correlogram of response 'Suicide' and numeric features", 
        ggplot2::aes(color = Suicide),
        progress = FALSE, 
        lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.1))) 
```

Generally, having attempted suicide (`Suicide` = "1") does not appear strongly related to any of these features in this dataset. There is a clear difference between the level distributions of `MD.TOTAL`, with the distribution for `Suicide` = "1" shifted higher than its counterpart for `Suicide` = "0". The former's median is roughly equal to the 75th percentile for the latter. Regarding collinearity, overall correlations are essentially non-existent between `Age` and each of `ADHD.Total` and `MD.TOTAL`, though the level-specific correlation values for `Suicide` = "1" are roughly twice the magnitude of those for `Suicide` = "0". The overall correlation between `ADHD.Total` and `MD.TOTAL` is approximately 0.482, suggesting the two features move in somewhat similar directions. Here again, there is a difference by level of `Suicide`: the value among individuals not having attempted suicide is approximately 0.562 compared with approximately 0.41 among individuals who have attempted it.


```{r suicide_feature1, fig.height=5.5}
df %>%
  filter(!is.na(Suicide)) %>%
  mutate(Suicide = fct_recode(Suicide,
                     'No attempt'='0',
                     Attempt='1'),
         Court.order = fct_recode(Court.order,
                                  No='0',
                                  Yes='1'),
         Hx.of.Violence = fct_recode(Hx.of.Violence,
                                     No='0',
                                     Yes='1'),
         Disorderly.Conduct = fct_recode(Disorderly.Conduct,
                                         No='0',
                                         Yes='1')) %>%
  ggpairs(columns = c('Suicide','Court.order','Hx.of.Violence','Disorderly.Conduct'), title = "Bar plots of response 'Suicide' and selected categorical features", 
        ggplot2::aes(color = Suicide),
        progress = FALSE,
        lower = list(discrete = "blank")) 
```

There could be differences in proportion between levels of `Suicide` relative to `Court.order`, `Hx.of.Violence`, and `Disorderly.Conduct`. Both `Court.order` and `Hx.of.Violence` skew towards "No"--that is, no court order and no history of violence--but individuals having attempted suicide (`Suicide` = "1") appear to show "Yes" in greater proportions than their non-attempting counterparts. By contrast, most individuals in the dataset have some history of disorderly conduct (`Disorderly.Conduct` = "1"), and the proportions appear relatively similar across levels of `Suicide`.


```{r suicide_feature2, fig.height=5.5}
df %>%
  filter(!is.na(Suicide)) %>%
  mutate(Suicide = fct_recode(Suicide,
                     'No attempt'='0',
                     Attempt='1'),
         Non.subst.Dx = fct_recode(Non.subst.Dx,
                                  'NA'='NA',
                                  None='0',
                                  One='1',
                                  'Two or more'='2'),
         Subst.Dx = fct_recode(Subst.Dx,
                               'NA'='NA',
                               None='0',
                               One='1',
                               Two='2',
                               'Three or more'='3'),
         Psych.meds. = fct_recode(Psych.meds.,
                                  'NA'='NA',
                                  None='0',
                                  One='1',
                                  'Two or more'='2')) %>%
  ggpairs(columns = c('Suicide','Non.subst.Dx','Subst.Dx','Psych.meds.'), title = "Bar plots of response 'Suicide' and selected categorical features", 
        ggplot2::aes(color = Suicide),
        progress = FALSE,
        lower = list(discrete = "blank")) 
```

Distributions for `Non.subst.Dx`, `Subst.Dx`, and `Psych.meds.` are relatively similar regardless of level of `Suicide`. Most individuals in the dataset show zero use of non-substance-related drugs, with fewer individuals at greater levels of use. Use of a single substance-related drug (`Subst.Dx` = "1") is most prevalent in the dataset, followed by zero use and use of two. As noted previously, `Psych.meds.` is primarily missing, and in general, individuals missing values for `Non.subst.Dx` and `Subst.Dx` are missing a value for `Psych.meds.`.


```{r suicide_feature3, fig.height=5.5}
df %>%
  filter(!is.na(Suicide)) %>%
  mutate(Suicide = fct_recode(Suicide,
                     'No attempt'='0',
                     Attempt='1'),
         Alcohol = fct_recode(Alcohol,
                              'NA'='NA',
                              'No use'='0',
                              Use='1',
                              Abuse='2',
                              Dependence='3'),
         THC = fct_recode(THC,
                          'NA'='NA',
                          'No use'='0',
                          Use='1',
                          Abuse='2',
                          Dependence='3'),
         Cocaine = fct_recode(Cocaine,
                              'NA'='NA',
                              'No use'='0',
                              Use='1',
                              Abuse='2',
                              Dependence='3'),
         Stimulants = fct_recode(Stimulants,
                                 'NA'='NA',
                                 'No use'='0',
                                 Use='1',
                                 Abuse='2',
                                 Dependence='3'),
         Sedative.hypnotics = fct_recode(Sedative.hypnotics,
                                         'NA'='NA',
                                         'No use'='0',
                                         Use='1',
                                         Abuse='2',
                                         Dependence='3'),
         Opioids = fct_recode(Opioids,
                              'NA'='NA',
                              'No use'='0',
                              Use='1',
                              Abuse='2',
                              Dependence='3')) %>%
  ggpairs(columns = c('Suicide','Alcohol','THC','Cocaine','Stimulants','Sedative.hypnotics','Opioids'), title = "Bar plots of response 'Suicide' and substance features", 
        ggplot2::aes(color = Suicide),
        progress = FALSE,
        upper = list(discrete = "blank"),
        lower = list(discrete = "blank")) 
```

Regarding the substance-misuse features, individuals in this dataset typically either show no use (\* = "0") or dependence (\* = "3"). Use (\* = "1") and abuse (\* = "2") are most common for `Alcohol`, `THC`, and `Cocaine`. Differences in proportions across levels of `Suicide` are most notable within the dependence group for `Alcohol`.

### Preprocessing

#### Imputation

#### Transformation


### 1. Clustering

#### Sub

### 2. Principal Component Analysis (PCA)

#### Sub

### 3. Support Vector Machine (SVM)

#### Sub


### Works Cited


### Code Appendix

The code chunks below represent the R code called in order during the analysis. They are reproduced in the appendix for review and comment.

```{r appendix, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

```{r loadpeng}
```
